The Ranked Nearest Neighbour Interchange space

### 2019-06-25 paper.tex
minor on definitions

A \amph{rooted binary phylogenetic tree} a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent?}
A \emph{ranked tree} is given by a pair of such a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function that maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The number $n$ of taxa will remain fixed throughout the whole paper.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge}.
The set of taxa that are descending from an internal edge $e$ is called \emph{cluster}. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes.
The set of clusters as defined above would then be recalled as non-trivial clusters.
Given a subset $M$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(M)$ of $M$ is the internal node with lowest rank that is ancestor of all taxa in $M$. \todo{OK?}
For a single taxon $x$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

The set of ranked trees builds the vertex set of the \emph{$\rnni$-graph}.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in that graph by an edge:
The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
Trees that are connected by either a rank swap or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space}.

@@

The main purpose of this paper is understanding geometric properties of the $\rnni$ space.
Therefore, further notations need to be introduced.
If every internal node of a tree $T$ has at least one leaf as child, $T$ is called \emph{caterpillar tree}.
A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T_1, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
For such a path $p$ we say the \emph{length} $|p|$ of $p$ is $m$.
A path $p$ between trees $T$ and $R$ is a \emph{shortest path} if for every path $r$ between $T$ and $R$, $|p| \leq |r|$.
The \emph{distance} $d(T,R)$ between trees $T$ and $R$ is the length of a shortest path between them.
If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths.
The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.
A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.

>>>>

A \amph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge}.
Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

Now we are ready to define the tree space that is the subject of study of this paper, the $\rnni$ graph.
The vertex set of the \emph{$\rnni$-graph} is given by the set of all ranked trees on $n$ taxa.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space}.

@@

The main purpose of this paper is understanding geometric properties of the $\rnni$ space.
Therefore, further notations need to be introduced.
If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called \emph{caterpillar tree}.
A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T_1, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
For such a path $p$ we say the \emph{length} $|p|$ of $p$ is $m$.
A path $p$ between trees $T$ and $R$ is a \emph{shortest path} if for every path $r$ between $T$ and $R$, $|p| \leq |r|$.
The \emph{distance} $d(T,R)$ between trees $T$ and $R$ is the length of a shortest path between them.
If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths between $T$ and $R$.
The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.
A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.
